\documentclass{article}
\usepackage{spconf,amsmath,graphicx,booktabs}
\usepackage{url}

\title{SpeechMaster: Confidence-Routed Self-Supervised ASR with Discrete Unit Budgeting}
\name{Zihang Zhou$^{\star\dagger}$ \qquad Kewei Liu$^{\star\dagger}$ \qquad Zihao Long$^{\star\dagger}$}
\address{$^{\star}$Shanghai Jiao Tong University \qquad $^{\dagger}$Equal contribution}

\begin{document}
\maketitle

\begin{abstract}
Self-supervised speech encoders are strong ASR front ends, but deploying the
largest recognizer on every utterance wastes computation and obscures the
accuracy/representation-cost tradeoff. We present SpeechMaster, a budget-aware
ASR framework that composes heterogeneous self-supervised representations. A
fast Wav2Vec2 CTC branch decodes every utterance, a label-free CTC risk score
routes uncertain utterances to a stronger HuBERT branch, and a HuBERT unit
auditor measures token rate and bitrate under discrete codebooks. On 1024
LibriSpeech dev-clean utterances, routing only 25\% of utterances reduces WER
from 2.94\% to 2.40\%, while full HuBERT decoding reaches 1.82\%. On 1024
test-clean utterances, the same router improves WER from 3.61\% to 3.04\% at a
25\% routing budget and to 2.60\% at a 50\% budget. An oracle route gives
1.45\% dev WER and 1.92\% test WER, showing complementary SSL-ASR errors.
\end{abstract}

\begin{keywords}
ASR, self-supervised speech, uncertainty routing, HuBERT, Wav2Vec2
\end{keywords}

\section{Introduction}

Low-resource ASR benefits from self-supervised learning (SSL) because encoders
such as Wav2Vec2, HuBERT, and WavLM learn phonetic and lexical structure from
unlabeled audio. A common downstream design selects one SSL checkpoint, attaches
or reuses a CTC decoder, and reports WER. This single-model view misses a
practical question: many utterances are easy for a fast recognizer, while a
smaller subset needs a stronger model. A useful low-resource system should
therefore expose an accuracy/compute knob instead of committing to one encoder.

SpeechMaster treats SSL-ASR as a routed composition. It always runs a fast
Wav2Vec2 branch, estimates utterance risk from CTC posterior geometry, and
spends the HuBERT large branch only on utterances selected by a budget. In
parallel, a discrete-unit auditor clusters HuBERT hidden states to quantify the
bitrate needed by unit-based downstream variants. The contributions are:
(1) a confidence-routed SSL-ASR framework, (2) a label-free CTC risk score with
budgeted inference, (3) a discrete-unit bitrate auditor over HuBERT layers and
codebook sizes, and (4) reproducible dev/test experiments with WER, CER, RTF,
token rate, bitrate, and routing ablations.

\section{SpeechMaster}

\subsection{Two-Branch SSL-ASR}

Given waveform $x$, a pretrained SSL encoder $f_\theta$ produces contextual
hidden states $H=f_\theta(x)$. A CTC projection maps each frame to token
posteriors:
\begin{equation}
    p_\theta(y_t \mid x)=\mathrm{softmax}(W H_t+b).
\end{equation}
SpeechMaster uses Wav2Vec2 base as the fast branch $F$ and HuBERT large as the
strong branch $S$. Both are public SSL-ASR checkpoints; SpeechMaster itself
adds no extra supervised training.

\subsection{CTC Risk Router}

Let $q_t$ denote the fast-branch posterior and let $\mathcal{T}$ be non-blank
argmax frames. SpeechMaster computes a risk score from posterior peakiness and
entropy:
\begin{equation}
    c(x)=\frac{1}{|\mathcal{T}|}\sum_{t\in\mathcal{T}}\max_k q_t(k)
    - \lambda \frac{1}{|\mathcal{T}|}\sum_{t\in\mathcal{T}}H(q_t),
\end{equation}
where $\lambda=0.03$. For routing budget $B$, utterances with the lowest
$c(x)$ are decoded by $S$; all others keep the fast hypothesis:
\begin{equation}
    \hat{y}(x)=
    \begin{cases}
    S(x), & x \in \mathrm{BottomB}(c),\\
    F(x), & \mathrm{otherwise}.
    \end{cases}
\end{equation}
The same saved branch outputs are reused to ablate peak-only, entropy-only,
duration, random, and oracle routers.

\subsection{Discrete Unit Auditor}

To connect ASR accuracy with representation compression, SpeechMaster extracts
HuBERT hidden states from several layers and trains MiniBatchKMeans codebooks.
For codebook size $K$ and token rate $r_{\mathrm{tok}}$, the nominal bitrate is
\begin{equation}
    \mathrm{bitrate}=r_{\mathrm{tok}}\log_2K.
\end{equation}
We also collapse consecutive duplicate units, giving a compact token stream
useful for text-to-unit or unit-to-ASR analyses.

\section{Experimental Setup}

Experiments use LibriSpeech clean splits at 16 kHz. The main development suite
uses 1024 dev-clean utterances; the final validation suite uses 1024 test-clean
utterances. Unit analysis uses 512 dev-clean utterances. The fast branch is
Wav2Vec2-Base-960h; the strong branch is HuBERT-Large-960h; a stronger
reference model is Wav2Vec2-Large-LV60. HuBERT unit analysis uses HuBERT-Base.
Exact checkpoint identifiers are listed in the released scripts. Metrics are
WER, CER, and real-time factor (RTF) for ASR, plus token rate, deduplicated
token rate, bitrate, and codebook size for discrete units.

\section{Results}

\subsection{Main Dev-Clean Results}

Table~\ref{tab:dev_route} is the main SpeechMaster development result. At a
25\% routing budget, WER drops from 2.94\% to 2.40\%, a relative reduction of
18.5\%, while using substantially less strong-branch decoding than 100\%
HuBERT. At 50\%, WER reaches 2.04\%. The oracle route is lower than both
branches on dev-clean, showing that fast and strong SSL recognizers make
non-identical errors.

\begin{table}[t]
\centering
\caption{SpeechMaster routing on 1024 LibriSpeech dev-clean utterances.}
\label{tab:dev_route}
\small
\resizebox{\columnwidth}{!}{%
\IfFileExists{../results/tables/heavy/speechmaster_metrics.tex}{\input{../results/tables/heavy/speechmaster_metrics.tex}}{
\begin{tabular}{lccc}
\toprule
Route & WER(\%) & CER(\%) & RTF\\
\midrule
0 & -- & -- & --\\
25 & -- & -- & --\\
100 & -- & -- & --\\
\bottomrule
\end{tabular}}}
\end{table}

Table~\ref{tab:asr} reports single-model and layer ablations. The final
Wav2Vec2 CTC layer is stronger than last-layer averaging or mid/final blending,
indicating that the pretrained CTC head is calibrated to the final state. The
large LV60 Wav2Vec2 model is the best single model on this dev slice, but the
fast+HuBERT oracle still improves over it.

\begin{table}[t]
\centering
\caption{Single-model ASR and layer ablations on dev-clean.}
\label{tab:asr}
\small
\resizebox{\columnwidth}{!}{%
\IfFileExists{../results/tables/heavy/asr_metrics.tex}{\input{../results/tables/heavy/asr_metrics.tex}}{
\begin{tabular}{lccc}
\toprule
System & WER(\%) & CER(\%) & RTF\\
\midrule
Wav2Vec2 final & -- & -- & --\\
HuBERT final & -- & -- & --\\
\bottomrule
\end{tabular}}}
\end{table}

\subsection{Test-Clean Verification}

Table~\ref{tab:test_route} verifies the same framework on test-clean. The
absolute WERs are higher than dev-clean, but the monotonic routing behavior
remains: 25\% routing improves WER from 3.61\% to 3.04\%, and 50\% routing
reaches 2.60\%. The oracle route reaches 1.92\%, confirming remaining
headroom for better risk prediction.

\begin{table}[t]
\centering
\caption{SpeechMaster routing on 1024 LibriSpeech test-clean utterances.}
\label{tab:test_route}
\small
\resizebox{\columnwidth}{!}{%
\IfFileExists{../results/tables/test_clean/speechmaster_metrics.tex}{\input{../results/tables/test_clean/speechmaster_metrics.tex}}{
\begin{tabular}{lccc}
\toprule
Route & WER(\%) & CER(\%) & RTF\\
\midrule
0 & -- & -- & --\\
25 & -- & -- & --\\
100 & -- & -- & --\\
\bottomrule
\end{tabular}}}
\end{table}

\subsection{Router and Unit Ablations}

Routing-score ablations show that risk-aware ordering is consistently better
than random routing. On dev-clean at 25\%, random routing gives 2.66\% WER,
duration gives 2.54\%, entropy gives 2.31\%, and SpeechMaster peak-entropy gives
2.40\%; the oracle gives 1.45\%. On test-clean at 50\%, random gives 2.93\%,
duration gives 2.62\%, peak-only gives 2.59\%, and peak-entropy gives 2.60\%.
Thus the proposed score is robust, but the oracle gap indicates that a learned
router is a promising extension.

The unit auditor exposes a compression tradeoff. Raw HuBERT assignments appear
at about 49.9 tokens/s for all tested layers. Deduplication roughly halves the
rate: for example, layer $-4$ with $K=100$ yields 27.14 deduplicated tokens/s
and 180.34 bit/s, while $K=1000$ raises this to 29.74 tokens/s and
296.36 bit/s. Layer $-8$ keeps more rapidly changing units than layer $-4$ at
the same codebook size, suggesting a higher acoustic-detail bitrate.

\begin{table}[t]
\centering
\caption{Selected HuBERT unit statistics on 512 dev-clean utterances.}
\label{tab:units}
\small
\begin{tabular}{lrrrr}
\toprule
Layer & $K$ & Tok/s & Dedup tok/s & Dedup bit/s\\
\midrule
$-8$ & 100 & 49.90 & 28.08 & 186.58\\
$-8$ & 1000 & 49.90 & 35.17 & 350.46\\
$-4$ & 100 & 49.90 & 27.14 & 180.34\\
$-4$ & 1000 & 49.90 & 29.74 & 296.36\\
$-1$ & 100 & 49.90 & 26.67 & 177.19\\
$-1$ & 1000 & 49.90 & 34.50 & 343.85\\
\bottomrule
\end{tabular}
\end{table}

\section{Conclusion}

SpeechMaster combines fast and strong self-supervised ASR models through
label-free CTC risk routing and audits the bitrate of HuBERT discrete units.
Across dev-clean and test-clean, routing a limited fraction of utterances gives
a smooth WER/RTF tradeoff, while oracle routing reveals substantial
complementarity. The results support a system-level view of low-resource
SSL-ASR: encoder choice, routing budget, and representation bitrate should be
optimized jointly rather than treated as isolated design choices.

\section{References}

\begin{thebibliography}{9}
\bibitem{librispeech}
V. Panayotov, G. Chen, D. Povey, and S. Khudanpur, ``Librispeech: An ASR
corpus based on public domain audio books,'' in \emph{Proc. ICASSP}, 2015.

\bibitem{wav2vec2}
A. Baevski, H. Zhou, A. Mohamed, and M. Auli, ``wav2vec 2.0: A framework for
self-supervised learning of speech representations,'' in \emph{NeurIPS}, 2020.

\bibitem{hubert}
W.-N. Hsu et al., ``HuBERT: Self-supervised speech representation learning by
masked prediction of hidden units,'' \emph{IEEE/ACM TASLP}, 2021.

\bibitem{wavlm}
S. Chen et al., ``WavLM: Large-scale self-supervised pre-training for full stack
speech processing,'' \emph{IEEE JSTSP}, 2022.

\bibitem{jiwer}
Jitsi, ``JiWER: Word error rate and character error rate for automatic speech
recognition,'' \url{https://jitsi.github.io/jiwer/}.
\end{thebibliography}

\end{document}
